The Weakly Nonlinear Magnetorotational Instability in a Global, Cylindrical Taylor-Couette Flow

### 2016-10-05 paper/widegap_mri_paper.tex
added james's edits

The magnetorotational instability (MRI) is widely belived to drive angular momentum transport in astrophysical disks. The MRI is a local instability excited by weak magnetic fields in differentially rotating fluids, and since first applied to an astrophysical context \citep{Balbus:1991vs} it has been widely invoked to explain accretion in protoplanetary disks \citep{Armitage:2010} and disks around black holes \citep{Blaes:2014}, as well as jet and wind launching (\citei{Lesur:2013}), anisotropic turbulence \citep{Murphy:2015tn}, and dynamo generation \citep{Brandenburg:1995, Vishniac:2009il}.

@@

The diversity of astrophysical systems which may be MRI unstable yields an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI--and even its very existence--may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, and so forth. Multiphysics numerical simulations of such systems are currently an area of intense focus, enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study \citep[e.g.][among many others]{Fleming:2003fs,Bai:2011cm, Flock:2013,Suzuki:2014vh}. Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. For example, the magnetic Prandtl number $\Pm = \nu/\eta \sim 10^{-8}$ in protoplanetary disks \citep[e.g.][]{Oishi:2011ei} and $\sim 10^{-6}$ in liquid metal experiments \citep[e.g.][]{Goodman:2002ix}. Such extreme ratios of viscosity to resistivity far exceed current computational resources. However, we can construct asymptotic approximations valid for $\Pm \ll 1$ using analytic methods.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_16E-7}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [ -8E-4, -1.3E-4, 1.3E-4, 8E-4, 1.5E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 3.31$ and $k_c = 0.902$ correspond to the smallest parameter values that yield a zero growth rate.}\label{fig:growth_rates}
\end{figure}

@@

Thus through judicious choice of cylinder rotation rates, one can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that the narrow gap approximation imposes a linear shear (constant $q$), and thus the interaction of fluid perturbations with the base velocity profile differs significantly from the case considered here. Our base velocity is

@@

We note that Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are written in a nonstandard form, with the nonlinear terms on the righthand side. This choice has a practical motivation. As detailed in \S\ref{sec:wnl_analysis}, we expand these equations in a perturbation series and solve them order by order using a pseudospectral code. The code solves partial differential equations of the form $\mathrm{M} \partial_t \mathbf{V} + \mathrm{L} \mathbf{V} = \mathbf{F}$, where $\mathrm{M}$ and $\mathrm{L}$ are matrices and $\mathbf{F}$ is a vector containing any inhomogenous terms. The nonlinear terms in our perturbation analysis become inhomogenous term inputs to the solver.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_totalstress_intstress.png}
\caption{Total stress $\mathrm{T}(r, z) = u_r u_\phi - \frac{2}{\beta} B_r B_\phi$ , i.e. the sum of the Reynolds and Maxwell stresses for the fiducial standard MRI parameters (top panel). Bottom panel shows the vertically integrated stress $\int \mathrm{T}(r, z) \, dz$.}\label{fig:total_stress}
\end{figure}

@@

We examine the behavior of the wide gap MRI system as a function of $\Pm$ in the regime $\Pm \ll 1$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $\Pm$, as well as the GLE linear coefficient $b$ and the diffusion coefficient $d$. From Equation \ref{eq:gle} it is readily apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, and we plot the dependence of $\alpha_{s}$ on $\Pm$ in the bottom panel of \ref{fig:coefficients}. 
Note that because $\reym$ is essentially constant as a function of $\Pm$, the saturation amplitude is equivalently sensitive to $\reye^{-1}$. We find by fitting the data that the saturation amplitude scales as $\alpha_{s} \sim \Pm^{0.777}$. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $\Pm^{2/3}$. They find that this amplitude dependence is driven by the $\Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in TC flow (\citei{Goodman:2002ix}).
Figure \ref{fig:thin_wide_comparison} shows the structure of the third-order nonlinear term $N_{31}^{(A)}$ as a function of $\Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$, and thus the scaling of the saturation amplitude because of the insensitivity of $b$ to $\Pm$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs} nad Paper I for the narrow gap equations). Clearly, the width of the boundary layers scales with $\Pm$ in both the wide and narrow gap MRI. This translates to a steeper saturation amplitude $\Pm$ dependence in the wide gap case.

@@

The phase dynamics of the complex GLE are well-studied in a variety of systems, and depend on the values of the GLE coefficients. The complex GLE may be unstable to traveling wave instabilities such as the Benjamin-Feir instability, a generalization of the Ekhaus instability. The complex GLE can also admit spatiotemporal chaos, and various classes of coherent structures (\citei{Aranson:2002}).Although a detailed description of the phase dynamics in the helical MRI is beyond the scope of this work, we note that such long-wavelength, long-timescale behavior may be observed in liquid metal helical MRI experiments.

@@

This is the first weakly nonlinear analysis of the MRI in a cylindrical geometry, and is thus the global analogue of similar analyses in local approximations \citep{Umurhan:2007hs,Vasil:2015}. Understanding the connection between local and global MRI modes is crucial for interpreting simulation results across domain geometries. (The relationship between local and global \textit{linear} MRI modes is investigated in \citet{Latter:2015}.) Phenomena such as saturation and the development of turbulence depend critically on the nature of the underlying MRI modes. The formalism presented here describes analytically the weakly nonlinear behavior of global MRI modes. This treatment should be expanded to encompass more astrophysically relevant conditions, so that our understanding of complicated MRI phenomena may continue to make contact with analytical theory.

@@

To find a bounded solution at $\mathcal{O}(\epsilon^3)$ we must eliminate secular terms: terms which are resonant with the solution to the linear homogenous equation $(\mathcal{L} + \xi \widetilde{\mathcal{H}}) \mathbf{V} = 0$ and cause the solution to grow without bound. Secular terms in our system are those that are resonant with the linear ansatz (Equation \ref{eq:ansatz}), i.e. terms with $e^{i k_c z}$ z-dependence. To eliminate these terms we enforce a solvability condition, which arises from a corollary to the Fredholm alternative. The Fredholm alternative states that if we consider a system of equations $\mathcal{L} \mathbf{V} = \mathbf{b}$ and its adjoint homogenous system $\mathcal{L}^\dagger \mathbf{V}^\dagger = 0$, only one of two conditions holds. Either there exists one and only one solution to the inhomogenous system, or the homogenous adjoint equation has a nontrivial solution. The relevant corollary arises as a consequence of the second condition: if $\mathcal{L}^\dagger \mathbf{V}^\dagger = 0$ has a nontrivial solution, then $\mathcal{L} \mathbf{V} = \mathbf{b}$ has a solution if and only if $\langle \mathbf{V}^\dagger | b \rangle = 0$.

>>>>

The magnetorotational instability (MRI) is believed to drive angular momentum transport in astrophysical disks. The MRI is a local instability excited by weak magnetic fields in differentially rotating fluids, and since first applied to an astrophysical context \citep{Balbus:1991vs} it has been invoked to explain accretion in protoplanetary disks \citep{Armitage:2010} and disks around black holes \citep{Blaes:2014}, as well as jet and wind launching (\citei{Lesur:2013}), anisotropic turbulence \citep{Murphy:2015tn}, and dynamo generation \citep{Brandenburg:1995, Vishniac:2009il}.

@@

The diversity of astrophysical systems which may be MRI unstable yields an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI -- and even its very existence -- may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, and so forth. Multiphysics numerical simulations of such systems are currently an area of intense focus, enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study \citep[e.g.][among many others]{Fleming:2003fs,Bai:2011cm, Flock:2013,Suzuki:2014vh}. Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. For example, the magnetic Prandtl number $\Pm = \nu/\eta \sim 10^{-8}$ in protoplanetary disks \citep[e.g.][]{Oishi:2011ei} and $\sim 10^{-6}$ in liquid metal experiments \citep[e.g.][]{Goodman:2002ix}. Such extreme ratios of viscosity to resistivity far exceed current computational resources. However, we can construct asymptotic approximations valid for $\Pm \ll 1$ using analytic methods.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_16E-7}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [-8E-4, -1.3E-4, 1.3E-4, 8E-4, 1.5E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 3.31$ and $k_c = 0.902$ correspond to the smallest parameter values that yield a zero growth rate.}\label{fig:growth_rates}
\end{figure}

@@

Thus through judicious choice of cylinder rotation rates, one can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that the narrow gap approximation imposes a linear shear (constant $q$), and so the interaction of fluid perturbations with the base velocity profile differs significantly from the case considered here. Our base velocity is

@@

We note that Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are written in a nonstandard form, with the nonlinear terms on the righthand side. This choice has a practical motivation. As detailed in \S\ref{sec:wnl_analysis}, we expand these equations in a perturbation series and solve them order by order using a pseudospectral code. The code solves partial differential equations of the form $\mathrm{M} \partial_t \mathbf{V} + \mathrm{L} \mathbf{V} = \mathbf{F}$, where $\mathrm{M}$ and $\mathrm{L}$ are matrices and $\mathbf{F}$ is a vector containing any nonhomogenous terms. The nonlinear terms in our perturbation analysis become nonhomogenous term inputs to the solver.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_totalstress_intstress.png}
\caption{Total stress $\mathrm{T}(r, z) = u_r u_\phi - \frac{2}{\beta} B_r B_\phi$, i.e. the sum of the Reynolds and Maxwell stresses for the fiducial standard MRI parameters (top panel). Bottom panel shows the vertically integrated stress $\int \mathrm{T}(r, z) \, dz$.}\label{fig:total_stress}
\end{figure}

@@

We examine the behavior of the wide gap MRI system as a function of $\Pm$ in the regime $\Pm \ll 1$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $\Pm$, as well as the GLE linear coefficient $b$ and the diffusion coefficient $d$. From Equation \ref{eq:gle} it is apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, and we plot the dependence of $\alpha_{s}$ on $\Pm$ in the bottom panel of Figure \ref{fig:coefficients}. 
Note that because $\reym$ is essentially constant as a function of $\Pm$, the saturation amplitude is equivalently sensitive to $\reye^{-1}$. We find by fitting the data that the saturation amplitude scales as $\alpha_{s} \sim \Pm^{0.777}$. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $\Pm^{2/3}$. They find that this amplitude dependence is driven by the $\Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in TC flow (\citei{Goodman:2002ix}).
Figure \ref{fig:thin_wide_comparison} shows the structure of the third-order nonlinear term $N_{31}^{(A)}$ as a function of $\Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$, and thus the scaling of the saturation amplitude because of the insensitivity of $b$ to $\Pm$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs} and Paper I for the narrow gap equations). Clearly, the width of the boundary layers scales with $\Pm$ in both the wide and narrow gap MRI. This translates to a steeper saturation amplitude $\Pm$ dependence in the wide gap case.

@@

The phase dynamics of the complex GLE are well-studied in a variety of systems, and depend on the values of the GLE coefficients. The complex GLE may be unstable to traveling wave instabilities such as the Benjamin-Feir instability, a generalization of the Ekhaus instability. The complex GLE can also admit spatiotemporal chaos, and various classes of coherent structures (\citei{Aranson:2002}). Although a detailed description of the phase dynamics in the helical MRI is beyond the scope of this work, we note that such long-wavelength, long-timescale behavior may be observed in liquid metal helical MRI experiments.

@@

This is the first weakly nonlinear analysis of the MRI in a cylindrical geometry, and is thus the global analogue of similar analyses in local approximations \citep{Umurhan:2007hs,Vasil:2015}. Understanding the connection between local and global MRI modes is crucial for interpreting simulation results across domain geometries. (The relationship between local and global \textit{linear} MRI modes is investigated in \citeauthor{Latter:2015} \citeyear{Latter:2015}.) Phenomena such as saturation and the development of turbulence depend critically on the nature of the underlying MRI modes. The formalism presented here describes analytically the weakly nonlinear behavior of global MRI modes. This treatment should be expanded to encompass more astrophysically relevant conditions, so that our understanding of complicated MRI phenomena may continue to make contact with analytical theory.

@@

To find a bounded solution at $\mathcal{O}(\epsilon^3)$ we must eliminate secular terms: terms which are resonant with the solution to the linear homogenous equation $(\mathcal{L} + \xi \widetilde{\mathcal{H}}) \mathbf{V} = 0$ and cause the solution to grow without bound. Secular terms in our system are those that are resonant with the linear ansatz (Equation \ref{eq:ansatz}), i.e. terms with $e^{i k_c z}$ z-dependence. To eliminate these terms we enforce a solvability condition, which arises from a corollary to the Fredholm alternative. The Fredholm alternative states that if we consider a system of equations $\mathcal{L} \mathbf{V} = \mathbf{b}$ and its adjoint homogenous system $\mathcal{L}^\dagger \mathbf{V}^\dagger = 0$, only one of two conditions holds. Either there exists one and only one solution to the nonhomogenous system, or the homogenous adjoint equation has a nontrivial solution. The relevant corollary arises as a consequence of the second condition: if $\mathcal{L}^\dagger \mathbf{V}^\dagger = 0$ has a nontrivial solution, then $\mathcal{L} \mathbf{V} = \mathbf{b}$ has a solution if and only if $\langle \mathbf{V}^\dagger | b \rangle = 0$.
